---
title: "Demonstration of Functions for Ecobat "
author: "Steve Markham"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---


```{r, echo = TRUE, warning = FALSE, message = FALSE}
#Libraries used
library(tidyverse) # Enables a consistent approach to data science
library(StreamMetabolism) # for sunset / sunrise times
library(gridExtra) # plots on a grid
library(broman) #helpful functions
library(data.table) # to write CSV
library(knitr) # Controls the R-code of Markdwon documents
library(hms) # Handles time
library(kableExtra) # Makes nice tobles
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}
#print out code by default
opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
# Function files 

source("convert-to-scientific-or-common.R")
# Convert_to_Sci()
# Bat_Sci_Com()

source("read-bat-data.R")
# read_batexplorer()
# read_kaleidoscope()
# Check_Static_Column()
# is_date_ymd

source("utility.R")
# Fm_date()
# Bat_Pallet()
# SpeciesColourScale()
# get_legend()

source("static-graph-functions.R")
# ActivePlot2017a()
# Static_outGraphs()
# Static_legend()
# NrGraph_String()
# Matrix_String()
# Static_Nightly_Activity()

source("exit-reentry-graph.R")
# NightData()
# Exit_Reentry_Times()
# Exit_Reentry_Graph()

```

# Data (pro forma) 

The structure of the data file required to produce the summary tables and info graphics.

```{r echo = FALSE, warning = FALSE, message = FALSE}

DataTable <- readxl::read_excel("Data//DataFormatTable.xlsx", 
                             sheet = 1, 
                             col_names = TRUE)


DataTable3 <- DataTable %>% 
   select(-Example)
```

```{r nice-tab12, tidy=FALSE, echo = FALSE, warning = FALSE, message = FALSE}  

knitr::kable(DataTable, "html", booktabs = TRUE, caption = 'Structure of Bat Data') %>% 
  kable_styling(bootstrap_options = "striped", full_width = F, font_size = 12) %>% 
  column_spec(1, bold = T, border_right = T) %>% 
  column_spec(2, bold = T, border_right = T, monospace = T) %>% 
  column_spec(3, border_right = T) %>% 
  column_spec(4, italic = T) # %>% 
```

# Read  _csv_ files (BatLogger and Kalidascope)  

## Bat Logger

The function `read_batexplorer` reads an `.csv` file exported from Elekon BatExplorer software <www.batlogger.com> and generates a `tibble` (data.frame) with the recommended data structure for bat analysis given above. The structure of `batexplorer-example.csv` a file exported from Elekon BatExplorer software is shown below:  

```{r echo = FALSE, warning = FALSE, message = FALSE}
#csv file exported from BatExplorer software

csv_file <- readr::read_csv("Data//batexplorer-example.csv", col_names = T) 

glimpse(csv_file)

```

The r chunk below reads the exported transect survey data `batexplorer-example.csv` for a site following sound analysis with _BatExplorer_. The _metadata_, key information that makes the data useful, not included with the exported file from _BatExplore_ is added via a character vector `Metadata`. 

```{r}
# function to convert csv file exported from BatExplorer to the "pro forma" 
# MetaData Vector - key information that makes the data useful
MetaData <- c("Site B",        # [1] Site Name
              "Transect 3",           # [2] Detector Location
              "Steve Markham",        # [3] Consultant
              "Elekon",               # [4] Detector Make
              "Batlogger M",          # [5] Detector Model
              "BatExplorer",          # [6] Bat Explorer
              "Version 1.11.4.0",     # [7] Software Version
              "Pass 2s gap",          # [8] Pass Definition
              "Manual ID",            # [9] Classification Method
              "Do not publish",       # [10] Data Sensitivity
              ""                      # [11] Any Notes
              )  
# read the csv and save to a data.frame pro_forma_data
# "Data//batexplorer-example.csv" is Location of expoted 
# BatExplorer csv path/filename.csv
pro_forma_data <- read_batexplorer("Data//batexplorer-example.csv", MetaData)  
```

The `batexplorer-example.csv` converted to _pro forma_ structure: 

```{r echo = FALSE, warning = FALSE, message = FALSE}
# Look at the data.frame
glimpse(pro_forma_data)
```


## Kaleidoscope

The function `read_kaleidoscope` reads an `.csv` file exported from Wildlife Acoustics Kaleidoscope software <https://www.wildlifeacoustics.com/products/kaleidoscope-software-ultrasonic> and generates a `tibble` (data.frame) with the recommended data structure for bat analysis given above. The structure of `kaleidoscope-example.csv` a file exported from Wildlife Acoustics Kaleidoscope software is shown below:  

```{r echo = FALSE, warning = FALSE, message = FALSE}
#csv file exported from BatExplorer software

csv_file <- readr::read_csv("Data//kaleidoscope-example.csv", col_names = T) 

glimpse(csv_file)

```

The r chunk below reads the exported transect survey data `kaleidoscope-example.csv` for a site following sound analysis with _Kaleidoscope_. The _metadata_, key information that makes the data useful, not included with the exported file from _Kaleidoscope_ is added via a character vector `Metadata`; including longitude and latitude as it is not automatically recorded. 

```{r}

# For Kaleidoscope Pro file 
MetaData <- c("Beaulieu Park",        # [1] Site Name
              "Transect 3",           # [2] Detector Location
              "Steve Markham",        # [3] Consultant
              "Wildlife Acoustics",   # [4] Detector Make
              "SM4BAT FS",            # [5] Detector Model
              "Kaleidoscope Pro",     # [6] Bat Explorer
              "Version",              # [7] Software Version
              "Pass 2s gap",          # [8] Pass Definition
              "Manual ID",            # [9] Classification Method
              "Do not publish",       # [10] Data Sensitivity
              ""                      # [11] Any Notes
              ) 
# SM4 does not give Long and Lat automatically so needs to be provided
Long <- 0.551
Lat <- 51.880
pro_forma_data <- read_kaleidoscope("Data//kaleidoscope-example.csv", 
                                    MetaData, 
                                    long, lat)
```

The `kaleidoscope-example.csv` converted to _pro forma_ structure: 

```{r echo = FALSE, warning = FALSE, message = FALSE}
# Look at the data.frame
glimpse(pro_forma_data)
```


# Summary Tables 

To Create Table and Graphs a `Night` variable is needed. The  original `pro_forma_data` to `Site_Data`; so it can be manipulated.

```{r}
#Transfer original pro_forma_data to Site_Data; so it can be manipulated
Site_Data <- pro_forma_data

#To Create Table and Graphs a Night variable is needed
Site_Data <- Site_Data %>% 
  mutate (Night = DateTime - lubridate::hours(12), #DateTime minus 12 hours
          Night = lubridate::as_date(Night)) #make Night variable just a date
```


## Table of Observations by Night

The code below produces a table of observations from the `Site_Data` data.frame. To help understanding, the code is reproduced in full (it could be made a function); the comments help explain the algorithm.


```{r, tidy=FALSE}

# Sum bat observations "activity" by night from Site_Data
BatTable <- Site_Data %>% 
  #remove the NAs
  filter(!is.na(ScientificName), !is.na(Night)) %>% 
  #group species into the night they were observed
  group_by(ScientificName, Night) %>% 
  # count the number of observations
  count(ScientificName) %>%  
  # order highest number first
  arrange(desc(Night)) 

#How many nights
NightNr <- length(levels(factor(BatTable$Night)))

#Make a Common Name Column and move to column 1
BatTable <- BatTable %>%
  #ungrouped so date format can be changed
  ungroup(Night) %>% 
  mutate(Common = Bat_Sci_Com(ScientificName),
         # Make a more readable date
         Night = Fm_date(Night, 1)) %>%
  select(Night, Common, everything())

# Make the columns names you want
colnames(BatTable) <- c("Night", "Common Name", "Scientific Name", "Observations (Nr)")

# This code makes the Table below
knitr::kable(BatTable, "html", booktabs = TRUE, 
             caption = stringr::str_c('Species observations over ',
                                      as.character(NightNr),
                                      " nights")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), 
                full_width = F, 
                font_size = 12, 
                position = "left") %>% 
  row_spec(0, bold = T, color = "white", background = "black") %>% 
  column_spec(2, italic = T) 

```

# Visualisation  

```{r echo = FALSE, warning = FALSE, message = FALSE}
#remove the NA's so plots don't have NA displayed
Site_Data <- Site_Data %>% 
  filter(!is.na(ScientificName))
```


## Time Zone and Daylight Saving

The imported data may not have the correct time zone or have been adjusted for daylight saving, the ability to correct this in the code is desirable; the R-code below adjusts the time, minus one hour, and sets the `timezone` as `Europe/London`. 

* `timezone` Local time should be used in the UK; this should follow the _iana_ Time Zone reference (tz) _Europe/London_ it takes into account daylight saving.  

```{r}
#Make Time Zone Local and make correct time 
Site_Data$DateTime <- lubridate::with_tz(Site_Data$DateTime, "Europe/London")
Site_Data$DateTime <- Site_Data$DateTime - lubridate::hours(1) 
```


## Nightly Activity from Statics

The function that generates static detector graph is `Static_Nightly_Activity`. The following Chunk calls the `Static_Nightly_Activity` with:

* the filtered data `StaticData` used to plot the graph  
* Alpha is the transparency of the bat activity spot (0 - no colour : 1 - full colour); Alpha can be turned down for numerous observations     
* panel.fill is the graph background colour `"midnightblue"`. An alternative `NULL` = none


```{r nice-fig81, warning = FALSE, message = FALSE, fig.cap='Example Static Activity plot', out.width='80%', fig.asp=1.1, fig.align='center'}

# Transfer Site_Data to a StaticData data.frame
StaticData <- Site_Data

# Visualisation uses Sc
StaticData$Scientific <- StaticData$ScientificName

# Plot graph
Static_Nightly_Activity(StaticData, 
                        Alpha = 0.8, 
                        panel.fill <- "midnightblue")

```

## Emergence Graphic

The Emergence & Re-entry Graphics are created by the function `Exit_Reentry_Graph` . 


```{r}
# Plot graph
Exit_Reentry_Graph(StaticData, Dusk = TRUE, Alpha = 0.4)

```


The above R Chunck uses the `GraphData` (data.frame) to plot the individual bat emergence times relative to Sunset:

* `Dusk` is set to `TRUE` to plot emergence activity;  
* `Alpha` (`dot transparancy`) is set to `0.3`(30%) to help distinguish the high level of activity.  

The plotted data could also be that observed on a _Transect_ or a group of _static detectors_. 

## Re-entry Graphic

Figure below shows the dawn re-entry graph note that the argument list has:

* `Dusk` is set to `FALSE` to plot reentry activity;  


```{r }
# Plot graph
Exit_Reentry_Graph(StaticData, Dusk = FALSE, Alpha = 0.5)

```

The `Exit_Reentry_Graph` calculates the sun times and the bat activity relative to the sun times using a couple functions `NightData` and `Exit_Reentry_Times`:   

* `NightData` uses the bat data (e.g. `GraphData`) and the function `sunrise.set` from the  `StreamMetabolism` package a table (`tibble`) is created of sunset, sunrise and night length from the `Night` range and location (`latitude` and `longitude`)    
* `Exit_Reentry_Times` merges, by `Night`, the table generated in `NightData` with the observation data (in the example above `GraphData`) and calculates the time each bat was observed relative to sunset and sunrise.  